---
title: "Week8"
format: html
---


```{r}
library(fpp3)
gafa_stock |>
    filter(Symbol == "AMZN")|>
    mutate(t = row_number())|>
    update_tsibble(index = t)|>
    gg_tsdisplay(Close, plot_type = "partial")
```


3.For the following series, find an appropriate Box-Cox transformation and order of differencing in order to obtain stationary data.

Turkish GDP from global_economy.
Accommodation takings in the state of Tasmania from aus_accommodation.
Monthly sales from souvenirs.

```{r}
# --- 1. Turkish GDP ---
gdp_tr <- global_economy |> filter(Country == "Turkey")
gdp_tr |> autoplot(GDP) + labs(title = "Turkish GDP")

# Box-Cox lambda (Guerrero) and required differencing
gdp_tr |> features(GDP, guerrero)
gdp_tr |> features(GDP, list(unitroot_ndiffs, unitroot_nsdiffs))

# Check the chosen transformation is now stationary
gdp_tr |>
  mutate(GDP_bc = box_cox(GDP, 0.157)) |>
  gg_tsdisplay(difference(GDP_bc), plot_type = "partial") +
  labs(title = "Turkish GDP: Box-Cox(0.157) + 1 difference")
```

**Turkish GDP.** Guerrero's method suggests $\lambda = 0.157$; a power transformation is
needed because GDP grows multiplicatively. The series is annual, so there is no
seasonal differencing (`nsdiffs = 0`). One non-seasonal difference (`ndiffs = 1`)
makes the series stationary. The ACF/PACF of the transformed, differenced series
show no remaining autocorrelation structure.

```{r}
# --- 2. Accommodation takings, Tasmania ---
tas <- aus_accommodation |> filter(State == "Tasmania")
tas |> autoplot(Takings) + labs(title = "Tasmania accommodation takings")

tas |> features(Takings, guerrero)
tas |> features(Takings, list(unitroot_ndiffs, unitroot_nsdiffs))

tas |>
  mutate(Takings_bc = box_cox(Takings, 0.002)) |>
  gg_tsdisplay(difference(Takings_bc, 4) |> difference(), plot_type = "partial") +
  labs(title = "Tasmania: log + seasonal(4) + 1 difference")
```

**Accommodation takings, Tasmania.** Guerrero's method gives $\lambda \approx 0.002$,
so a log transformation is appropriate (the seasonal amplitude grows with the level).
The data are quarterly: one seasonal difference at lag 4 brings the seasonal pattern
under control, and one further non-seasonal difference is needed
(`ndiffs = 1`, `nsdiffs = 1`).

```{r}
# --- 3. Monthly sales, souvenirs ---
souvenirs |> autoplot(Sales) + labs(title = "Souvenir monthly sales")

souvenirs |> features(Sales, guerrero)
souvenirs |> features(Sales, list(unitroot_ndiffs, unitroot_nsdiffs))

souvenirs |>
  mutate(Sales_bc = box_cox(Sales, 0.002)) |>
  gg_tsdisplay(difference(Sales_bc, 12) |> difference(), plot_type = "partial") +
  labs(title = "Souvenirs: log + seasonal(12) + 1 difference")
```

**Souvenir sales.** Guerrero's method gives $\lambda \approx 0.002$, again indicating
a log transformation, since the seasonal variation increases strongly with the level.
The data are monthly, so one seasonal difference at lag 12 removes the seasonality,
followed by one non-seasonal difference (`ndiffs = 1`, `nsdiffs = 1`).

