---
title: "Week9"
format: html
---

```{r}
library(fpp3)
```

# Section 9.11 exercises (Q6–Q10)

## Q6. Simulating data from simple ARIMA models

```{r}
set.seed(2026)

# (a) AR(1) with phi_1 = 0.6, sigma^2 = 1
y <- numeric(100)
e <- rnorm(100)
for (i in 2:100) y[i] <- 0.6 * y[i - 1] + e[i]
sim <- tsibble(idx = seq_len(100), y = y, index = idx)
sim |> autoplot(y) + labs(title = "AR(1) with phi1 = 0.6")
```

**(b) Effect of $\phi_1$.** A helper that re-simulates with any $\phi_1$:

```{r}
sim_ar1 <- function(phi, seed) {
  set.seed(seed)
  ee <- rnorm(100); yy <- numeric(100)
  for (i in 2:100) yy[i] <- phi * yy[i - 1] + ee[i]
  tibble(idx = seq_len(100), y = yy, phi = paste0("phi1 = ", phi))
}
bind_rows(sim_ar1(-0.6, 1), sim_ar1(0, 2), sim_ar1(0.6, 3), sim_ar1(0.9, 4)) |>
  ggplot(aes(idx, y)) + geom_line() +
  facet_wrap(~phi, scales = "free_y") +
  labs(title = "AR(1): effect of phi1")
```

**Reading the plot.** With $\phi_1 = 0$ the series is white noise (no persistence). As $\phi_1$
grows towards $+1$ the series becomes increasingly smooth and wavy: neighbouring values are
strongly positively related, so runs above/below the mean last longer. With $\phi_1 = -0.6$ the
series oscillates rapidly, jumping from positive to negative values each period. Only $|\phi_1| < 1$
gives a stationary process; larger values (e.g. $\phi_1 = 1$) are non-stationary and drift.

```{r}
# (c) MA(1) with theta_1 = 0.6
set.seed(2026)
e <- rnorm(100)
y <- numeric(100)
for (i in 2:100) y[i] <- e[i] + 0.6 * e[i - 1]
sim_ma <- tsibble(idx = seq_len(100), y = y, index = idx)
sim_ma |> autoplot(y) + labs(title = "MA(1) with theta1 = 0.6")
```

**(d) Effect of $\theta_1$.**

```{r}
sim_ma1 <- function(theta, seed) {
  set.seed(seed)
  ee <- rnorm(100); yy <- numeric(100)
  for (i in 2:100) yy[i] <- ee[i] + theta * ee[i - 1]
  tibble(idx = seq_len(100), y = yy, theta = paste0("theta1 = ", theta))
}
bind_rows(sim_ma1(-0.6, 5), sim_ma1(0, 6), sim_ma1(0.6, 7), sim_ma1(0.9, 8)) |>
  ggplot(aes(idx, y)) + geom_line() +
  facet_wrap(~theta, scales = "free_y") +
  labs(title = "MA(1): effect of theta1")
```

**Reading the plot.** An MA(1) has memory of only one period, so the series always looks like
noisy white noise; changing $\theta_1$ does not create visible smoothness the way $\phi_1$ does.
What changes is the lag-1 autocorrelation (with $\theta_1 = 0$ it is plain white noise, and with
large $|\theta_1|$ consecutive observations become mildly related). Unlike an AR(1), any value of
$\theta_1$ still gives a stationary series.

```{r}
# (e) ARMA(1,1) with phi_1 = 0.6, theta_1 = 0.6
set.seed(2026)
e <- rnorm(100); y <- numeric(100)
for (i in 2:100) y[i] <- 0.6 * y[i - 1] + e[i] + 0.6 * e[i - 1]
arma11 <- tsibble(idx = seq_len(100), y = y, index = idx) |> mutate(model = "ARMA(1,1)")

# (f) AR(2) with phi_1 = -0.8, phi_2 = 0.3 (non-stationary)
set.seed(2026)
e <- rnorm(100); y <- numeric(100)
for (i in 3:100) y[i] <- -0.8 * y[i - 1] + 0.3 * y[i - 2] + e[i]
ar2 <- tsibble(idx = seq_len(100), y = y, index = idx) |> mutate(model = "AR(2)")
```

**(g) Comparing the two series.**

```{r}
arma11 |> as_tibble() |>
  bind_rows(ar2 |> as_tibble()) |>
  ggplot(aes(idx, y)) + geom_line() +
  facet_wrap(~model, scales = "free_y") +
  labs(title = "ARMA(1,1) vs AR(2)")
```

**Comparison.** The ARMA(1,1) with $\phi_1 = \theta_1 = 0.6$ is stationary: it wanders around zero
with moderate persistence, because the AR and MA parts partly offset each other. The AR(2) with
$\phi_1 = -0.8$, $\phi_2 = 0.3$ is *non-stationary* — the AR polynomial has roots inside the unit
circle — so the simulated series shows increasing, explosive oscillations whose amplitude grows
with time.

## Q7. `aus_airpassengers`

**(a) ARIMA model, diagnostics and forecasts.**

```{r}
fit <- aus_airpassengers |> model(ARIMA(Passengers))
report(fit)
fit |> gg_tsresiduals()
```

```{r}
fit |> forecast(h = 10) |> autoplot(aus_airpassengers) +
  labs(title = "aus_airpassengers: ARIMA forecasts (10 years)")
augment(fit) |> features(.innov, ljung_box, lag = 10)
```

**Result.** `ARIMA()` selects an **ARIMA(0,2,1)** model with $\hat\theta_1 = -0.8963$
(AICc = 198.3). Because the series is trended and grows multiplicatively, two non-seasonal
differences are used. The Ljung–Box test gives $p = 0.75$, so there is no evidence against the
residuals being white noise (the residual ACF and histogram also look clean). The forecasts
continue the upward, slightly accelerating trend.

**(b) Backshift-operator form.** Using $\hat\theta_1 = -0.8963$,

$$(1-B)^2 y_t = (1 + \theta_1 B)\varepsilon_t = (1 - 0.8963B)\varepsilon_t,$$

i.e. $y_t - 2y_{t-1} + y_{t-2} = \varepsilon_t - 0.8963\,\varepsilon_{t-1}$.

**(c) ARIMA(0,1,0) with drift.**

```{r}
fit_c <- aus_airpassengers |>
  model(`ARIMA(0,1,0)+drift` = ARIMA(Passengers ~ pdq(0,1,0) + 1))
report(fit_c)
fit_c |> forecast(h = 10) |>
  autoplot(aus_airpassengers) +
  labs(title = "ARIMA(0,1,0) w/ drift forecasts")
```

This is a random walk with drift: the forecasts are a straight line with slope
$\hat c = 1.42$ million passengers per year. They are similar to part (a) in the short run but are
**linear**, whereas the ARIMA(0,2,1) forecasts are slightly curved and grow a little faster. The
AICc (200.6) is barely worse than the selected model.

**(d) ARIMA(2,1,2) with drift.**

```{r}
fit_d  <- aus_airpassengers |> model(`ARIMA(2,1,2)+drift` = ARIMA(Passengers ~ pdq(2,1,2) + 1))
fit_d0 <- aus_airpassengers |> model(`ARIMA(2,1,2) no const` = ARIMA(Passengers ~ pdq(2,1,2) + 0))
report(fit_d)
report(fit_d0)
```

The ARIMA(2,1,2) with drift fits slightly better in-sample ($\hat\sigma^2 = 4.03$) but uses five
parameters, giving a worse AICc (206.6) than the parsimonious models. Its forecasts are very
similar to (a) and (c). **Removing the constant** makes the model fail: `ARIMA()` reports
"non-stationary AR part from CSS" and returns a NULL model, because an ARIMA(2,1,2) without a
constant cannot capture the growing trend.

**(e) ARIMA(0,2,1) with a constant.**

```{r}
fit_e <- aus_airpassengers |> model(`ARIMA(0,2,1)+const` = ARIMA(Passengers ~ pdq(0,2,1) + 1))
report(fit_e)
```

`ARIMA()` warns that "the model specification induces a quadratic or higher order polynomial
trend". Adding a constant to an already twice-differenced model introduces a deterministic
quadratic trend, so the forecasts bend upwards implausibly quickly. This is why the constant is
normally dropped from an ARIMA(0,2,1).

## Q8. United States GDP (`global_economy`)

```{r}
us_gdp <- global_economy |> filter(Country == "United States")
us_gdp |> autoplot(GDP) + labs(title = "US GDP")
us_gdp |> features(GDP, guerrero)
us_gdp |> features(GDP, unitroot_ndiffs)
```

**(a) Transformation.** Guerrero's method gives $\lambda = 0.28$, so a power transformation
(approximately a cube-root / $\lambda \approx 0.3$) is appropriate — the series grows
multiplicatively and its variability increases with the level. `unitroot_ndiffs` suggests
$d = 2$, but the automatic model selection below settles on $d = 1$ with drift, which behaves
similarly.

**(b) Automatic ARIMA on the transformed data.**

```{r}
fit_gdp <- us_gdp |> model(ARIMA(box_cox(GDP, 0.3)))
report(fit_gdp)
fit_gdp |> gg_tsresiduals()
augment(fit_gdp) |> features(.innov, ljung_box, lag = 10)
```

`ARIMA()` selects an **ARIMA(1,1,0) with drift** on the $\lambda = 0.3$ data:
$\hat\phi_1 = 0.46$ and drift $\hat c = 200.5$. The Ljung–Box test gives $p = 0.96$ — the
residuals behave like white noise.

**(c)–(d) Trying other orders and choosing a model.**

```{r}
gdp_fits <- us_gdp |>
  model(
    `ARIMA(1,1,0)+drift` = ARIMA(box_cox(GDP, 0.3) ~ pdq(1,1,0) + 1),
    `ARIMA(0,2,1)`       = ARIMA(box_cox(GDP, 0.3) ~ pdq(0,2,1)),
    `ARIMA(0,2,2)`       = ARIMA(box_cox(GDP, 0.3) ~ pdq(0,2,2)),
    `ARIMA(2,2,1)`       = ARIMA(box_cox(GDP, 0.3) ~ pdq(2,2,1)),
    ARIMA                = ARIMA(box_cox(GDP, 0.3))
  )
gdp_fits |> glance() |> arrange(AICc) |> select(.model, sigma2, AIC, AICc, BIC)
```

```{r}
fit_best <- us_gdp |> model(`ARIMA(0,2,2)` = ARIMA(box_cox(GDP, 0.3) ~ pdq(0,2,2)))
report(fit_best)
fit_best |> gg_tsresiduals()
augment(fit_best) |> features(.innov, ljung_box, lag = 10)
```

**Best model.** By AICc the **ARIMA(0,2,2)** is preferred
($\hat\theta_1 = -0.50$, $\hat\theta_2 = -0.25$, AICc = 708.4), just ahead of ARIMA(0,2,1) and
ARIMA(2,2,1). Its Ljung–Box $p = 0.95$ and residual plots are satisfactory. (The automatic
ARIMA(1,1,0)+drift is a close, more parsimonious alternative.)

**(e) Forecasts.**

```{r}
fit_best |> forecast(h = 20) |>
  autoplot(us_gdp |> filter(Year >= 2000) |> mutate(GDP = box_cox(GDP, 0.3)), level = 90) +
  labs(title = "US GDP (Box-Cox 0.3) forecasts", y = "box_cox(GDP, 0.3)")
```

The forecasts are reasonable: they continue the upward trend of the transformed GDP, and the
prediction intervals widen slowly (much more slowly than on the original scale would suggest,
which is the point of the transformation).

**(f) Comparison with `ETS` (no transformation).**

```{r}
cmp <- us_gdp |> model(ARIMA = ARIMA(box_cox(GDP, 0.3)), ETS = ETS(GDP))
cmp |> accuracy() |> select(.model, RMSE, MAE, MAPE)
cmp |> forecast(h = 20) |>
  autoplot(us_gdp |> filter(Year >= 2000), level = 90) +
  labs(title = "US GDP: ARIMA (Box-Cox) vs ETS forecasts")
```

Both models capture the trend well and give similar point forecasts. On the training data the
ARIMA (Box-Cox) model has slightly smaller RMSE/MAE/MAPE. The main practical difference is the
prediction intervals: because ETS is fitted to the raw (untransformed) GDP, its intervals widen
much faster, so its long-horizon forecasts are far less precise than the ARIMA-on-log forecasts.

## Q9. `aus_arrivals`

**(a) Selecting a country — the UK.**

```{r}
uk <- aus_arrivals |> filter(Origin == "UK")
uk |> autoplot(Arrivals) + labs(title = "UK arrivals to Australia")
```

The time plot shows a clear **upward trend** over 1981–2012 with a **strong, stable quarterly
seasonal pattern** (peaks in the December and March quarters, troughs mid-year). The seasonal
amplitude grows as the level rises, suggesting a multiplicative (or log) structure.

**(b) Differencing to stationarity.**

```{r}
uk |> features(Arrivals, list(unitroot_nsdiffs, unitroot_ndiffs))
uk |> mutate(d = difference(Arrivals, 4) |> difference()) |>
  gg_tsdisplay(d, plot_type = "partial") +
  labs(title = "UK arrivals: seasonal(4) + first difference")
```

Both a **seasonal difference** ($\text{nsdiffs} = 1$, at lag 4) and a **non-seasonal difference**
($\text{ndiffs} = 1$) are needed, i.e. $(1-B)(1-B^4)y_t$.

**(c) ACF of the differenced series.**

```{r}
d <- uk |> mutate(d = difference(Arrivals, 4) |> difference()) |> filter(!is.na(d))
d |> ACF(d, lag_max = 12) |> as_tibble() |> print(n = 12)
```

The ACF has a **large negative spike at lag 1** ($\approx -0.60$), a **negative spike at the
seasonal lag 4** ($\approx -0.47$) and a positive spike at lag 5 ($\approx 0.50$), with the
remaining values decaying gradually. The absence of a sharp cut-off and the significant seasonal
lags point to both a non-seasonal and a seasonal MA component (rather than a pure MA(1) or a
seasonal-only model).

**(d) PACF of the differenced series.**

```{r}
d |> PACF(d, lag_max = 12) |> as_tibble() |> print(n = 12)
```

The PACF also shows large spikes at **lag 1** ($\approx -0.60$), **lag 2** ($\approx -0.41$) and
the **seasonal lag 4** ($\approx -0.44$), then decays. Spikes in both ACF and PACF are the
signature of a model containing **both AR and MA terms**, non-seasonal and seasonal.

**(e) Model suggested by the plots.** A mixed model such as
**ARIMA(1,1,1)(1,1,1)$_4$**, or more simply ARIMA(0,1,1)(0,1,1)$_4$, is suggested: the differencing
pushes the spikes to lag 1 and lag 4, and both AR and MA structure is present.

**(f) What does `ARIMA()` choose?**

```{r}
uk_fit <- uk |> model(ARIMA(Arrivals))
report(uk_fit)
uk_fit |> gg_tsresiduals()
augment(uk_fit) |> features(.innov, ljung_box, lag = 12)
```

```{r}
uk_fit |> forecast(h = 8) |> autoplot(uk, level = 90) +
  labs(title = "UK arrivals: selected ARIMA forecasts")
```

`ARIMA()` selects **ARIMA(2,0,2)(0,1,2)$_4$** — the same *kind* of model my ACF/PACF reading
suggested (mixed AR and MA, with seasonal MA terms), but with higher orders. The automatic
selection (chosen by AICc) is preferable here: the extra AR/MA terms are needed because the
one-lag spikes were not clean cut-offs, and its residuals pass the Ljung–Box test ($p = 0.25$).

**(g) Backshift-operator form.** Writing the selected model as

$$(1-\phi_1B-\phi_2B^2)(1-B^4)\,y_t = (1+\theta_1B+\theta_2B^2)(1+\Theta_1B^4+\Theta_2B^8)\,\varepsilon_t,$$

and substituting $\hat\phi_1 = 1.516$, $\hat\phi_2 = -0.592$,
$\hat\theta_1 = -1.419$, $\hat\theta_2 = 0.752$, $\hat\Theta_1 = -0.766$,
$\hat\Theta_2 = 0.523$:

$$(1-1.516B+0.592B^2)(1-B^4)\,y_t = (1-1.419B+0.752B^2)(1-0.766B^4+0.523B^8)\,\varepsilon_t .$$

Expanding the backshift polynomials gives the model without backshift operators.

## Q10. A series from `us_employment` — US Retail Trade

```{r}
ret <- us_employment |> filter(Title == "Retail Trade")
ret |> autoplot(Employed) + labs(title = "US Retail Trade employment")
```

**(a) STL decomposition.**

```{r}
ret |> model(STL(Employed ~ season(window = "periodic"))) |>
  components() |> autoplot() +
  labs(title = "STL decomposition: US Retail Trade")
```

The trend rises over the whole period but with clear slowdowns (e.g. around the early-1980s and
1990s recessions and the 2008 GFC). The seasonality is **strong and stable**, with a distinct
within-year pattern (a large build-up towards the December holiday season and a sharp January
drop) and the seasonal amplitude growing with the level of the series.

**(b) Transformation.**

```{r}
ret |> features(Employed, guerrero)
```

Guerrero's method gives $\lambda \approx 0.55$, so a transformation is warranted
(roughly a square root, $\lambda \approx 0.5$), because the seasonal swings widen as the level
grows.

**(c) Stationarity / differencing.**

```{r}
ret |> features(Employed, list(unitroot_nsdiffs, unitroot_ndiffs))
```

Both a **seasonal difference** (`nsdiffs = 1`) and a **non-seasonal difference**
(`ndiffs = 1`) are required after the transformation, i.e. $(1-B)(1-B^{12})y_t$.

**(d)–(e) Candidate models and the best by AICc.**

```{r}
ret_fits <- ret |>
  model(
    `ARIMA(0,1,1)(0,1,1)[12]` = ARIMA(box_cox(Employed, 0.5) ~ pdq(0,1,1) + PDQ(0,1,1)),
    `ARIMA(0,1,2)(0,1,1)[12]` = ARIMA(box_cox(Employed, 0.5) ~ pdq(0,1,2) + PDQ(0,1,1)),
    `ARIMA(2,1,0)(1,1,0)[12]` = ARIMA(box_cox(Employed, 0.5) ~ pdq(2,1,0) + PDQ(1,1,0)),
    ARIMA = ARIMA(box_cox(Employed, 0.5))
  )
ret_fits |> glance() |> arrange(AICc) |> select(.model, sigma2, AIC, AICc, BIC)
```

```{r}
report(ret_fits |> select(ARIMA))
ret_fits |> select(ARIMA) |> gg_tsresiduals()
augment(ret_fits |> select(ARIMA)) |> features(.innov, ljung_box, lag = 24)
```

`ARIMA()` selects **ARIMA(1,1,2)(2,1,1)$_{12}$** on the $\lambda = 0.5$ data, and it has the
lowest AICc (1042.6) of the models considered. Its residuals broadly look like white noise, but
the Ljung–Box test is marginal ($p \approx 0.01$ at lag 24) — a little structure remains. A
simpler ARIMA(0,1,1)(0,1,1)$_{12}$ (a "common candidate") fits much worse
(AICc = 1126, Ljung–Box rejects strongly), so it is *not* adequate here. The selected model is the
better choice.

**(f) Forecasting the next three years.**

```{r}
fc <- ret |> model(ARIMA = ARIMA(box_cox(Employed, 0.5))) |> forecast(h = 36)
fc |> autoplot(ret, level = 90) +
  labs(title = "US Retail Trade employment: ARIMA forecasts (3 years)")
```

```{r}
fc |> hilo(level = 90) |> as_tibble() |>
  mutate(width = `90%`$upper - `90%`$lower, rel = width / .mean,
         yr = as.integer(Month - yearmonth("2019 Sep")) %/% 12 + 1) |>
  group_by(yr) |> summarise(rel_width = round(mean(rel), 3))
```

The point forecasts continue the recent trend and reproduce the seasonal shape. Comparing with
the actual FRED series (BLS retail-trade employment), the very near-term forecasts are close, but
the 2020 COVID shock makes the multi-year forecasts (which of course know nothing about it) miss
by a wide margin.

**(g) How far ahead are the forecasts usable?** The 90% interval width, as a fraction of the
forecast level, grows roughly like: year 1 $\approx 3\%$, year 2 $\approx 9\%$, year 3
$\approx 15\%$. As a rough rule, forecasts are "sufficiently accurate" while that interval stays
within about 10% of the level — i.e. roughly **two to three years** ahead. Beyond that the bands
become so wide that the forecasts are of limited practical use.
